---
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r warning=FALSE}
library(melt)
library(reshape2)
library(psych)
library(ggplot2)
library(tree)
```


# Question 2: Data Pre-processing Steps

# Load the dataset
```{r}
bushfire <- read.csv("BushFireData.csv")
```
# Check the structure of the dataset to identify data types
```{r}
str(bushfire)
```
# Remove rows with missing values to ensure a clean dataset for analysis
```{r}
bushfire <- na.omit(bushfire)
```

# Convert categorical variables (Construction Quality and Insurance Coverage) to factors
```{r}
bushfire$Construction_Quality <- as.factor(bushfire$Construction_Quality)
```

```{r}
bushfire$Insurance_Coverage <- as.factor(bushfire$Insurance_Coverage)
```

# Set a seed for reproducibility and create a 70-30 train-test split
```{r}
set.seed(1)
```

```{r}
tr.id = sample(1:nrow(bushfire), round(nrow(bushfire) * 0.7, 0))
```

# Define the training data (70% of the dataset) and testing data (remaining 30%)
```{r}
training_data = bushfire[tr.id, ]  # Training data
testing_data = bushfire[-tr.id, ]  # Test data
```

# Check the dimensions of the training and testing datasets to confirm the split
```{r}
dim(training_data)
dim(testing_data)
```

# Display the structure of the training dataset after pre-processing
```{r}
str(training_data)
```
# QUESTION 3

#Distribution of Fire Intensity
```{r}
hist(bushfire$Fire_Intensity, main = "Distribution of Fire Intensity", xlab = "Fire Intensity (kW/m)", col = "#008080")

```


```{r}
hist(bushfire$Distance_from_Fire, main = "Distribution of Distance from Fire", xlab = "Distance from Fire (km)", col = "#008080")
```


```{r}
hist(bushfire$Property_Value, main = "Distribution of Property Value", xlab = "Property Value (x 10^5 AUD)", col = "#008080")

```


```{r}
hist(bushfire$Damage_Claims, main = "Distribution of Damage Claims", xlab = "Damage Claims (x 10^5 AUD)", col = "#008080")
```

#2 bar plots
```{r}
barplot(table(bushfire$Construction_Quality), main = "Construction Quality Distribution", xlab = "Construction Quality", col = c("red", "green"))

```


```{r}
barplot(table(bushfire$Insurance_Coverage), main = "Insurance Coverage Distribution", xlab = "Insurance Coverage", col = c("red", "green", "pink"))
```

# boxplots
```{r}
boxplot(Damage_Claims ~ Construction_Quality, data = bushfire, 
        main = "Damage Claims by Construction Quality", 
        xlab = "Construction Quality", ylab = "Damage Claims (x 10^5 AUD)", col = "#008080")
```


```{r}
boxplot(Damage_Claims ~ Insurance_Coverage, data = bushfire, 
        main = "Damage Claims by Insurance Coverage", 
        xlab = "Insurance Coverage", ylab = "Damage Claims (x 10^5 AUD)", col = "#008080")
```
# Scatter Plots

```{r}
plot(bushfire$Fire_Intensity, bushfire$Damage_Claims, 
     main = "Damage Claims vs Fire Intensity", 
     xlab = "Fire Intensity (kW/m)", ylab = "Damage Claims (x 10^5 AUD)", pch = 20, col = "#008080")
```


```{r}
plot(bushfire$Distance_from_Fire, bushfire$Damage_Claims, 
     main = "Damage Claims vs Distance from Fire", 
     xlab = "Distance from Fire (km)", ylab = "Damage Claims (x 10^5 AUD)", col = "#008080" )
```


```{r}
# Select only continuous variables
continuous_vars <- bushfire[, sapply(bushfire, is.numeric)]
cor_matrix <- cor(continuous_vars, use = "complete.obs")
print(cor_matrix)
```
```{r}
cor_matrix <- cor(continuous_vars, use = "complete.obs")
melted_cor_matrix <- melt(cor_matrix)
```

```{r}
ggplot(data = melted_cor_matrix, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  coord_fixed()
```


# QUESTION 4

#Initial Model Building

```{r}
initial_model <- lm(Damage_Claims ~ Fire_Intensity + Distance_from_Fire + 
                      Building_Age + Property_Value + Population_Density + 
                      Emergency_Response_Time + Mitigation_Measures + 
                      Wind_Speed + Humidity, data=training_data)
```


```{r}
summary(initial_model)
```
# Model Improvement
```{r}
interaction_model <- lm(Damage_Claims ~ Fire_Intensity * Distance_from_Fire
+ Building_Age + Property_Value + Population_Density + Emergency_Response_Time + Mitigation_Measures + Wind_Speed + Humidity, data=training_data)
```

```{r}
summary(interaction_model)
```

# Polynomial model
```{r}
polynomial_model <- lm(Damage_Claims ~ poly(Fire_Intensity, 2) + Distance_from_Fire +  Building_Age + Property_Value + Population_Density + 
Emergency_Response_Time + Mitigation_Measures + Wind_Speed + Humidity, data=training_data)
```

```{r}
summary(polynomial_model)
```
# Variable Selection Using Stepwise Regression
```{r}
library(MASS)
optimized_model <- stepAIC(initial_model, direction="both")
summary(optimized_model)
```
```{r}
plot(optimized_model)
```

# Model Evaluation with RMSE

```{r}
predictions <- predict(optimized_model, testing_data)
rmse <- sqrt(mean((testing_data$Damage_Claims - predictions)^2))
rmse
```


# QUESTION 5
```{r}
final_model <- lm(Damage_Claims ~ Fire_Intensity + Distance_from_Fire + Property_Value + Emergency_Response_Time + Mitigation_Measures + Humidity, data = training_data)

summary(final_model)
```


```{r}
predictions <- predict(final_model, testing_data)
final_rmse <- sqrt(mean((testing_data$Damage_Claims - predictions)^2))
cat("Final Model RMSE:", final_rmse, "\n")
```
# QUESTION 7
# Model building
```{r warning=FALSE}
tree_train <- tree(Damage_Claims~.,training_data)

```


```{r}
plot(tree_train)
text(tree_train, pretty = 0)
```


```{r}
summary(tree_train)
```

```{r warning=FALSE}
set.seed(12)
cv_regtree <- cv.tree(tree_train)
```


```{r}
cv_regtree
```


```{r}

plot(cv_regtree$size, cv_regtree$dev, type = "b" )
```


#Prune the best model with the lowest "dev"
```{r warning=FALSE}
# Prune to size 15 for minimal deviance
pruned_tree_9 <- prune.tree(tree_train, best = 9)
predictions_9 <- predict(pruned_tree_9, testing_data)
```

#Check the model Accuracy
```{r warning=FALSE}
MSE_9 <- mean((testing_data$Damage_Claims - predictions_9)^2)
RMSE_9 <- sqrt(MSE_9)
RMSE_9
```

# Plot the best model Regression Tree
```{r}
pruned_tree_9
```
```{r}
plot(pruned_tree_9)
text(pruned_tree_9, pretty = 0)
```

```{r}
plot(predictions_9, testing_data$Damage_Claims)
abline(0, 1)
```

# Lets check if the model is biased.
Residuals should be randomly scattered around zero; patterns may suggest areas where the model struggles.
```{r}
residuals <- testing_data$Damage_Claims - predictions_9
plot(predictions_9, 
     residuals, xlab = "Predicted", ylab = "Residuals", main = "Residual Plot")
abline(h = 0, col = "red")

```


# QUESTION 8
The model is size with 9;
```{r}
MSE <- mean((testing_data$Damage_Claims - predictions_9)^2)
RMSE <- sqrt(MSE_9)
RMSE
```

# Terninal Nodes:
```{r}
pruned_tree_9
```

# Interpretting the model with size 9 
```{r}
plot(pruned_tree_9)
text(pruned_tree_9, pretty = 0)
```

# QUESTION 9
# Best model for regression
```{r}
summary(final_model)
```

# Best model for Tree
```{r}
summary(pruned_tree_9)

```


# Check the model Performance of both models 
```{r}
LinearModel <- predict(final_model, testing_data)
linearMSE <-  mean((testing_data$Damage_Claims - LinearModel)^2)
linearRMSE <- sqrt(linearMSE)
linearRMSE
```

# Tree RMSE
```{r}
RMSE
```
# Linear Regression Plot
```{r}
plot(testing_data$Damage_Claims, LinearModel, main = "Linear Regression: Actual vs Predicted",
     xlab = "Actual Damage Claims", ylab = "Predicted Damage Claims", col = "blue", pch = 19)
abline(0, 1, col = "red")  # 45-degree line
```
# Pruned Tree Model Plot
```{r}
plot(testing_data$Damage_Claims, predictions_9, main = "Pruned Tree: Actual vs Predicted",
     xlab = "Actual Damage Claims", ylab = "Predicted Damage Claims", col = "brown", pch = 19)
abline(0, 1, col = "red")  # 45-degree line

```


#QUESTION 10
# Create the new binary target variable for classification
# 'NewDamageClaims' is 1 if 'Damage_Claims' is greater than 7.5, otherwise it's 0
```{r}
bushfire$NewDamageClaims <- ifelse(bushfire$Damage_Claims > 7.5, 1, 0)
```

# Remove unnecessary variables
```{r}
head(bushfire$NewDamageClaims)
```

```{r}
str(bushfire)
```





